### 2015-03-31 paper/chemokine_supplement_revised.tex
RSI Submission

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration (SUPPLEMENT)}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\label{eq:dde}
\end{equation}
\vspace{.05in}
}

@@

{\footnotesize
\begin{equation*}
\label{EqS1}
\begin{aligned}
\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
\dot{T} &= \rho T -\beta TV \\
\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
\dot{V} &= pI - \beta TV - \gamma (t) V \\
\gamma (t) &= \left\{ \begin{array}{rcl}
	1/\mbox{day} & \mbox{,}  & t < 5  \\
	3/\mbox{day} & \mbox{,} & t \geq 5  
	\end{array}\right. \\
\end{aligned}
\tag{S1}
\end{equation*}
}
\vspace{0.5in}

@@

The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.  Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time. Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection. Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$. Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells. Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection). The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million. The initial number of infected cells is assumed to be zero. Parameter values are listed in Table S1 in Text S1.  This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.  The final value was found to be 1,257 per hour.

@@

Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we hypothesize that it renders IP-10 ineffectual.  We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).  Because of this behavior IP-10 was not included in the aH5N1 model runs.

@@

We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection (Table 3).  Parameters are presented along with their range using the following format:  \\

@@

The parameters in the first grouping did not affect the outcome of the infection unless they were adjusted to values outside the realm of possibility.  Specifically, each parameter in this group seems to lie within a range that does not significantly affect the asymptotic behavior of the model.  We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{mostly stable} parameters.  The difference between the two being that \textit{mostly stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.

Of interest, this group can be mostly split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).  Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.  The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal Effects.

The stability of these parameters makes sense in the context of our model.  Chemokine exists in our model to provide a chemical gradient that T cells may follow to the focus of infection.  The total quantity of the chemokine in the lung does not have a strong effect on the location and size of the gradient.  Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells may follow.

@@

While we have claimed that these parameters are qualitatively stable within a certain range, many of the parameters were set to values that did lead to a divergent model behavior.  We have deemed these deviations acceptable on an individual basis as described below. \\

@@

We chose to limit the increase of the apoptosis to a factor of two because larger values would not be biologically relevant.  In the other direction we lowered the value all the way to zero.

Adjusting the apoptosis time causes the model results to diverge slightly in the sH1N1 strain.  Also of interest, reducing the apoptosis time to zero still did not allow the immune response to clear the pandemic infection.  As stated earlier, its inclusion in this group is already suspect and its effects are described in more detail in the results section of the main paper. \\

@@

We tested values up to two orders of magnitude larger and smaller than the default parameter value.  These extremes are not biologically plausible but were tested anyway.

Varying the chemokine decay rate results in stable model behavior for values one order of magnitude larger and smaller than the default.  A value two orders of magnitude smaller results in divergent behavior for both the aH5N1 strain and the sH1N1 strain and a value two orders of magnitude larger results in divergent behavior for only the sH1N1 strain.  This maximum decay rate used for all three strains corresponds to an implausible 18 second half-life.  The minimal value corresponds to a similarly implausible 50 hour half-life.  The fact that an extremely low decay rate can hinder clearance is interesting.  In this case, lack of decay allows the chemokine to diffuse homogeneously across the entire infection, removing the concentration gradient required by T cells to find the active areas of the infection.  This suggests that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.  Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection. \\

@@

The chemokine diffusion rate only diverges in the aH5N1 strain for the highest value.  If anything, our estimate for the diffusion rate is already large as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.  Thus the diffusion rate we used based on Stokes-Einstein and water viscosity was likely within a physiologic range. \\

@@

Chemokine secretion values differ between strains.  It is important to node that aH5N1 and sH1N1 show a threshold at the same concentration: near 1e-6 $pg/s\cdot cell$.  While this may seem contradictory, it is actually an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.  Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of divergence. \\

@@

A value of 1s was chosen as the minimum due to computational constraints, yet this value is likely already too small to be biologically plausible.  We extended the range to up to 600 times the default value to examine what happens when the circulation time is so long that the T cells are effectively removed from the simulation if they miss the focus of infection on their first pass through the lung.

T cell circulation times were tested over a very large range and does diverge for high values.  Classifying this parameter as \textit{mostly stable} may seem counterintuitive when examining the sensitivity figures.  It is important to note that in this case, we tested extra values on the higher end of the baseline than other plots.  While each strain does show divergence beyond a certain value, the intermediate values of 1s and 20s remain qualitatively similar to the main result, demonstrating a stable range surrounding the default value.  Because vascular and lymph circulation times are unknown, parameter values of up to 1 minute and 3 minutes in mice may be plausible and warrant further study. \\

\textbf{T Cell Expected Kill Time:} [ \textit{0m | 10m | 100m} ]  - The T cell expected kill time defines the T cell kill rate, the rate at which T cells probabilistically induce apoptosis of  virus-expressing epithelial cells in their immediate vicinity.  A lower kill time corresponds to a higher kill rate, and vice-versa.

We tested values only one order of magnitude higher than the default as it does not seem reasonable to expect a T cell to take over an hour to induce apoptosis of a virus-expressing cell.  On the other side, we allowed T cells to induce apoptosis instantly to see if any delay in the induction of apoptosis was responsible for the inability for the immune response to clear the pandemic infection.

@@

T cell decay parameters allow the model to diverge only in the most extreme cases.  Neither of these values are reasonable and the parameters show stable behavior otherwise.  While the default of 2 hours may seem short, it appears to determine similar behavior as models with larger values.

@@

The two parameters in this grouping do seem to affect the model's behavior up until the introduction of the T cell response, followed by convergence of infected cell numbers back to the same levels.  Interestingly, these are the two parameters that define the transition delay between the different stages of epithelial cell infection. Because these terms refer to the timing of when infected cells release the virus particles, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response. \\

@@

We tested values up to a factor of two larger and smaller than the original amount.

While a longer incubation time does not change the number of virions released by each infected cell, it does slow down the spread of the infection.  Conversely, a shorter incubation time allows the infection to spread more quickly by allowing the infected cell to release the new virus in to the system earlier.  Thus, changes in the incubation time do affect the infection profile before the introduction of T cells.  The infection dynamics change dramatically upon the arrival of T cells.  As discussed, one of the main factors in determining whether or not the infection is cleared is the amount of new virus produced per infected cell.  While this does depend somewhat on the overall size of the plaque, this value depends heavily on shutting down the expression of new virus, something T cells are unable to do during the cellular incubation phase as the virus-incubating cell has not expressed virus for T cells to detect.  Thus, assuming the original plaque has not grown so large that it can no longer be effectively covered by the T cell response, the overall course of the infection will be unaffected by the length of the incubation period. \\

@@

The parameter was increased up by a factor of 3 and down by a factor of 10.  In each case, the change is extreme enough such that the resulting values are not biologically likely.

@@

\subsubsection{Unstable Parameters}

This leaves five parameters shown that appear to directly affect the result of the infection: IgM strength, infection sensitivity, T cell secretion rate, viral decay, and viral diffusion.  By comparing the three strains of influenza, we also know that the viral secretion rate affects the result.  Of interest, nearly all of these parameters are directly related to the behavior of the virus.  Only the T cell secretion rate and IgM parameters are not.  \\

\textbf{IgM Strength:} [ \textit{1 | 10 | 1,000} ] - IgM strength simulates the presence of IgM at day 4 post-infection by increasing the decay rate of free virus particles by the given factor.

@@

While it is true that the extreme values have a large effect on the model behavior, it is more of interest that the intermediate values also seem to change the model behavior.  This shows that there is no stable area around the default as is seen in the \textit{mostly stable} parameters.  The higher the strength of this parameter, the less virus there is in the system.  Because the true effective strength of IgM is unknown in the context of this model, it is not a target of our investigation. \\

@@

Viral decay has the exact same effect as the IgM strength parameter.  Its value directly determines how much virus remains in the system over the course of the infection.  There is a strong relationship between the decay rate and the resulting infection profile across all values tested. \\

\textbf{Infection Sensitivity:} [ \textit{12 m/virion | 2 h/virion | 20 h/virion} ] - The infection sensitivity describes the ability of the virus to infect healthy cells.  A larger value corresponds to a virus particle needing less time to infect a nearby healthy epithelial cell.

@@

The strength of the infection sensitivity parameter is directly related to how much virus is present in the area of the healthy cells.  Thus, increasing its value by any factor is equivalent to increasing the amount of virus by the same factor.  Thus, it behaves similarly to the IgM and viral decay parameters as it directly determines the virulence of the virus. \\

@@

Shown elsewhere, the viral secretion rate has a strong effect on the outcome of the infection.  Similar to the previously discussed parameters, its value directly affects how much virus there is in the system.  Thus, it's affect is similar to the previous parameters. \\

\textbf{Viral Diffustion Rate:} [ \textit{3e-4 $\mu m^2/s$ | 3e-2 $\mu m^2/s$ | 3 $\mu m^2/s$} ] - The viral diffusion rate controls how quickly the virus spreads out over the monolayer.

@@

The viral diffusion rate does not change how much virus there is in the system.  Rather, it determines how fast the virus may spread across the alveoli.  While its mechanism is different, its effect may be even stronger than the previous parameters.  By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may out-pace the body's ability to generate a chemical gradient.  Thus, T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection. \\

\textbf{T Cell Secretion Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell secretion rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.

@@

The T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.  Thus, it is reasonable to assume that there is a threshold of T cell secretion beyond which the dynamics of the infection do not change.  This is consistent with our observations of T cell clumping in areas of high chemokine concentration.  Increasing the number of T cells in the system does not seem to help beyond a certain point because the T cells overlap in space and become redundant.  The rapid decline of the infection size for the maximum value of $3 \mu m^2/s$ is simply an artifact of the infection exhausting the entire model space by day 3.

@@

@@

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=\textwidth]{Figure_S3}
	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.}
	\label{fig:asensitivity}
\end{center}
\end{figure}

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=\textwidth]{Figure_S4}
	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.}
	\label{fig:ssensitivity}
\end{center}
\end{figure}

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=\textwidth]{Figure_S5}
	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.}
	\label{fig:psensitivity}
\end{center}
\end{figure}

@@

%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}

% \end{table}

>>>>

% Use lscape packages to allow for specific pages to be displayed as landscape
% \usepackage{lscape}
\usepackage{pdflscape}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung (SUPPLEMENT)}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Judy Cannon$^{3}$,
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico Health Sciences Center, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation*}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\tag{S1}
\label{eq:dde}
\end{equation*}
\vspace{.05in}
}

@@

{\footnotesize
\begin{equation*}
\label{eq:miao}
\begin{aligned}
\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
\dot{T} &= \rho T -\beta TV \\
\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
\dot{V} &= pI - \beta TV - \gamma (t) V \\
\gamma (t) &= \left\{ \begin{array}{rcl}
	1/\mbox{day} & \mbox{,}  & t < 5  \\
	3/\mbox{day} & \mbox{,} & t \geq 5  
	\end{array}\right. \\
\end{aligned}
\tag{S2}
\end{equation*}
}
\vspace{0.5in}

@@

The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.  Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time. Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection. Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$. Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells. Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection). The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million. The initial number of infected cells is assumed to be zero. Parameter values are listed in Table S3.  This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.  The final value was found to be 1,257 per hour.

@@

Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we consider that IP-10 production is markedly reduced in comparison with that of RANTES and this may lead to elevated RANTES secretion measured in aH5N1 \textit{in vitro} infection when compared to the other strains.  Thus, IP-10 production was not included in aH5N1 simulations.

%Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.

% we hypothesize that it renders IP-10 ineffectual.  We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).  Because of this behavior IP-10 was not included in the aH5N1 model runs.

% we consider that IP-10 production is markedly reduced in comparison with that of RANTES and this may lead to elevated RANTES secretion measured in aH5N1 \textit{in vitro} infection when compared to the other strains.  Thus, IP-10 production was not included in aH5N1 simulations.

@@

We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior (stable parameters), parameters that affect the model's quantitative peak infection size but do not affect final clearance, (peak change parameters), and those parameters that affect the final clearance of the infection (sensitive parameters) (Table 3).  Parameters are presented along with their range using the following format:  \\

@@

Stable parameters do not affect the outcome of the infection unless they are adjusted to values outside the realm of possibility.  Specifically, the plausible parameter values in this group are within a range that does not significantly affect the asymptotic behavior of the model.  We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{bounded stable} parameters.  \textit{Bounded stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.

Of interest, this group can be split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).  Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.  The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal Effects.

The stability of these parameters makes sense in the context of our model.  Chemokine exists in our model to provide a chemical gradient that T cells may follow to the focus of infection.  The total quantity of the chemokine in the lung does not have a strong effect on the location and size of the gradient.  Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells are able to follow.

@@

While we have claimed that these parameters are qualitatively stable within a certain range, many of the parameters were set to values that did lead to a divergent model behavior.  Testing these extreme values provides bracketing information regarding the range of stability of the parameter in question.  We have deemed these deviations acceptable on an individual basis as described below. \\

Compartmental modeling design suggests the removal of such stable parameters.  Due to the nature of the ABM, inclusion of these parameters is mechanistically necessary.  Thus, we include and parameterize them using the methods described in the main paper and Table 2.

@@

We chose to limit the increase of the apoptosis to a factor of two because larger values would not be biologically relevant.  We considered values as low as zero.

Adjusting the apoptosis time causes the model results to diverge slightly in the sH1N1 strain (Fig.~\ref{fig:ssensitivity}).  Also of interest, reducing the apoptosis time to zero still did not allow the immune response to clear the pandemic infection.  As stated earlier, its inclusion in this group is marginal and its effects are described in more detail in the results section of the main paper. \\

@@

We tested values up to two orders of magnitude larger and smaller than the default parameter value even though these extremes are not biologically plausible.

Varying the chemokine decay rate results in stable model behavior for values one order of magnitude larger and smaller than the default.  A value two orders of magnitude smaller results in divergent behavior for both the aH5N1 strain and the sH1N1 strain and a value two orders of magnitude larger results in divergent behavior for only the sH1N1 strain.  This maximum decay rate used for all three strains corresponds to an implausible 18 second half-life.  The minimal value corresponds to a similarly implausible 50 hour half-life.  The fact that an extremely low decay rate can hinder clearance is interesting.  In this case, lack of decay allows the chemokine to diffuse homogeneously across the entire infection, removing the concentration gradient required by T cells to find the active areas of the infection.  This suggests that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.  % Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection. \\

@@

The chemokine diffusion rate only diverges in the aH5N1 strain for the highest value.  Our estimate for the diffusion rate is already elevated as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.  \\

@@

Chemokine secretion values differ between strains.  It is important to node that aH5N1 and sH1N1 show a threshold at the same concentration: near 1e-6 $pg/s\cdot cell$.  This is an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.  Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of divergence. \\

@@

A value of 1 second was chosen as the minimum due to computational constraints, yet this value is likely already too small to be biologically plausible.  We extended the range to up to 600 times the default value to examine what happens when the circulation time is so long that the T cells are effectively removed from the simulation if they miss the focus of infection on their first pass through the lung.

T cell circulation times were tested over a very large range and does diverge for high values.  Classifying this parameter as \textit{bounded stable} may seem counterintuitive when examining the sensitivity figures.  It is important to note that in this case, we tested extra values on the higher end of the baseline than other plots.  While each strain does show divergence beyond a certain value, the intermediate values of 1 second and 20 seconds remain qualitatively similar to the main result, demonstrating a stable range surrounding the default value.  Because vascular and lymph circulation times are unknown, parameter values of up to 1 minute and 3 minutes in mice may be plausible and warrant further study. \\

\textbf{T Cell Expected Kill Time:} [ \textit{0m | 10m | 100m} ]  - The T cell expected kill time defines the rate at which T cells probabilistically induce apoptosis of  virus-expressing epithelial cells in their immediate vicinity.  A lower kill time corresponds to a higher kill rate.

We tested values only one order of magnitude higher than the default limiting the time to induce apoptosis of a virus-expressing cell to one hour.  On the other side, we allowed T cells to induce apoptosis instantly to see if any delay in the induction of apoptosis was responsible for the inability for the immune response to clear the pandemic infection.

@@

T cell decay parameters allow the model to diverge only in the most extreme cases.  Neither of these values are reasonable and the parameters show stable behavior otherwise.  While the default of 2 hours may seem short, it appears to generate similar behavior to models with larger values.

@@

Difference in peak only parameters affect the model's behavior up until the introduction of the T cell response, followed by a return to similar day 10 outcomes.  Interestingly, these are the two parameters that define the transition delay between the different stages of epithelial cell infection. Because these terms refer to the timing of when infected cells release the virus particles, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response. \\

@@

We tested values up to a factor of two larger and smaller than the original parameter setting.

While a longer incubation time does not change the number of virions released by each infected cell, it does slow down the spread of the infection.  Conversely, a shorter incubation time allows the infection to spread more quickly by allowing the infected cell to release the new virus in to the system earlier.  Thus, changes in the incubation time do affect the infection profile before the introduction of T cells.  The infection dynamics change dramatically upon the arrival of T cells.  One of the main factors in determining whether or not the infection is cleared is the amount of new virus produced per infected cell.  While this does depend somewhat on the overall size of the plaque, this value depends heavily on shutting down the expression of new virus, something T cells are unable to do during the cellular incubation phase as the virus-incubating cell has not expressed virus for T cells to detect.  Thus, assuming the original plaque has not grown so large that it can no longer be effectively covered by the T cell response, the overall course of the infection will be unaffected by the length of the incubation period. \\

@@

The parameter was increased up by a factor of 3 and down by a factor of 10.  In each case, the change is extreme enough that the resulting values are not biologically likely.

@@

\subsubsection{Sensitive Parameters}

Sensitive parameters directly affect the result of the infection.  There are five parameters classified as sensitive: viral response to IgM, infectivity, T cell production rate, viral decay, and viral diffusion.  By comparing the three strains of influenza, we also know that the viral secretion rate affects the result.  Of interest, a majority of these parameters are related to the behavior of the virus. The only sensitive parameter not related to the behavior or the virus is the T cell production rate.  \\

\textbf{Viral Response to IgM:} [ \textit{1 | 10 | 1,000} ] - Viral response to IgM simulates the presence of IgM at day 4 post-infection by increasing the decay rate of free virus particles by the given factor.

@@

While it is true that the extreme values have a large effect on the model behavior, it is more of interest that the intermediate values also seem to change the model behavior.  This shows that there is no stable area around the default as is seen in the \textit{bounded stable} parameters.  The higher the strength of this parameter, the less virus there is in the system.  Because the true effective strength of IgM is unknown in the context of this model, it is not a target of our investigation. \\

@@

Viral decay has the exact same effect as the viral response to IgM parameter.  Its value directly determines how much virus remains in the system over the course of the infection.  There is a strong relationship between the decay rate and the resulting infection profile across all values tested. \\

\textbf{Infectivity:} [ \textit{12 m/virion | 2 h/virion | 20 h/virion} ] - Infectivity describes the ability of the virus to infect healthy cells.  A larger value corresponds to a virus particle needing less time to infect a nearby healthy epithelial cell.

@@

The strength of the infectivity parameter is directly related to how much virus is present in the area of the healthy cells.  Thus, viral density is linearly proportional to infectivity.  Thus, it behaves similarly to the IgM and viral decay parameters as it directly determines the virulence of the virus. \\

@@

The viral secretion rate has a strong effect on the outcome of the infection.  Similar to the previously discussed parameters, its value directly affects how much virus there is in the system.  Thus, its effect is similar to the previous parameters. \\

\textbf{Viral Diffusion Rate:} [ \textit{3e-4 $\mu m^2/s$ | 3e-2 $\mu m^2/s$ | 3 $\mu m^2/s$} ] - The viral diffusion rate controls how quickly the virus spreads across the monolayer.

@@

The viral diffusion rate does not change how much virus there is in the system.  Rather, it determines how fast the virus may spread across the alveoli.  While its mechanism is different, its effect may be even stronger than the previous parameters.  Increasing the viral diffusion rate so that the virus diffuses faster than the chemokine (an unlikely scenario) allows the virus to out-pace the body's ability to establish a chemical gradient around the infected cells.  In this scenario T cells are constantly directed to locations behind the rapidly spreading viral cloud and would be unable to `get ahead' of the infection. \\

\textbf{T Cell Production Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell production rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.

@@

The T cell production rate does have a consistent response over its different values, but this effect is minimized above a certain rate.  Thus, it is reasonable to assume that there is a threshold of T cell production beyond which the dynamics of the infection do not change.  This is consistent with our observations of T cell clumping in areas of high chemokine concentration.  Increasing the number of T cells in the system does not control the infection beyond a certain point because the T cells overlap in space and become redundant.  The rapid decline of the infection size for the maximum value of $3 \mu m^2/s$ is simply an artifact of the infection exhausting the entire model space by day 3.

\subsection{Window of Control}

We predict viral control in the model by approximating $R_0$ and determining if the value is above or below a value of one.  $R_0$ is calculated as the number of virions produced by each cell multiplied by the number of cells infected by each virion.  The first half of this equation is the rate of viral production multiplied by the time the cell produces virus.  The second part of the equation is termed the viral efficiency rate, $E$.  The equation here is repeated from the main paper:

\begin{equation}
R_0 = p \times (T_{arr} + T_{kill} + T_{apop}) \times E
\end{equation}

In the main paper, $E(t)$ is approximated as 0.27, and this value is accurate enough to provide useful predictions.  Calculating the true value for $E(t)$ is more difficult as it depends on many factors, including the area of the plaque, the depth of the ring of infected (but not dead cells) around the plaque, the local viral density, and the arrival rate and effectiveness of T cells.  Each of these factors change during the infection.

To find values of $E(t)$ for each of the three strains we reran the model in the absence of T cells and calculated the amount of virus that infects cells, $I(t)$, and the amount that decays $D(t)$, at each time step.  The viral effectiveness rate is then calculated to be

\begin{equation}
E(t) = \frac{I(t)}{I(t) + D(t)}
\end{equation}

This approach accounts for the factors listed above and provides a unique measure of $E(t)$ at each point in time.  We focus on results in the model after day 5 p.i. corresponding to the arrival of T cells in the main model.  In each simulation $I(t)$ remains relatively constant, while $D(t)$ increases linearly with time.  We perform least-squared regressions on $I(t)$ and $D(t)$ for each strain and combine the results to give an estimate for $E(t)$ (Table~\ref{tab:effectiveness}).

We use the estimated values of $E(t)$ to calculate $R_0$ values for each strain in the absence of T cell mediated apoptosis.  Because all three $E(t)$ values all change by less than 10\% over the five day window, we estimate $R_0$ values using the average $E(t)$ value for each strain (Table~\ref{tab:effectiveness}).  The three strains show similar values for $R_0$, suggesting a saturation of viral effectiveness as the plaque grows.   This is not the case early in the infection as pH1N1 growth clearly outpaces sH1N1 which in turn expands more rapidly than aH5N1.  In the main model, the T cell response limits viral production time, which in turn increases the effectiveness of the virus that is produced.  Thus, it is reasonable to assume viral effectiveness remains high in the presence of the T cell response.  By accounting for basic spatial and temporal effects as described in the main paper, we approximate $E(t)$ as 0.27 over the course of the infection.

@@

\section*{Videos}

Videos can be found at the following web address: http://cs.unm.edu\textbackslash\%7Edrew/chemokine

@@

\pagebreak
\begin{landscape}

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=9in]{Figure_S3}
	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.  In each subgraph, the thicker purple line shows the results of the model using the default parameter value.}
	\label{fig:asensitivity}
\end{center}
\end{figure}

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=9in]{Figure_S4}
	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.  In each subgraph, the thicker purple line shows the results of the model using the default parameter value.  The divergence between Apoptosis Time values of 90m and 120m (upper-left plot) highlights an important threshold of infection control.}
	\label{fig:ssensitivity}
\end{center}
\end{figure}

\begin{figure}[ht!]
\begin{center}
	\includegraphics[width=9in]{Figure_S5}
	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.  In each subgraph, the thicker purple line shows the results of the model using the default parameter value.}
	\label{fig:psensitivity}
\end{center}
\end{figure}

\end{landscape}

%\begin{figure}[ht!]
%\begin{center}
%	\includegraphics[width=\textwidth]{Figure_S6}
%	\caption{\textbf{pH1N1 sensitivity analysis.} Model results and fits of the viral effectiveness rate, $E$.}
%	\label{fig:effectiveness}
%\end{center}
%\end{figure}

@@

%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}
